Cluster Difference Imaging Photometric Survey (CDIPS) II. TOI~837: A Young Validated Planet in IC~2602

### 2020-07-28 paper/ms.tex
update manuscript
Co-authored-by: Alan Levine <aml@space.mit.edu>

The observations of \tn\ were initially acquired by
\citet{ziegler_soar_2020} as part of the Southern Astrophysical
Research (SOAR) TESS Survey using the High Resolution Camera (HRCam;
\citealt{tokovinin_ten_2018}).  The HRCam $I$-band filter is described
by \citet{tokovinin_ten_2018}.  The points in
Figure~\ref{fig:followup} show the resulting measured 5-$\sigma$
detectable contrasts.  The lines are linear smoothing fits between the
regimes of the diffraction limit, the ``knee'' at $\approx 0.2''$, and
the slow decrease until $\approx 1.5''$, beyond which the speckle
patterns become de-correlated.  Star A (TIC 847769574) was detected at
the expected location and brightness contrast, and no additional
companions were found.  Star B was not detected; with a separation of
$5.4''$ from \tn, it fell outside the field of view.

@@

\begin{figure}[!t]
	\begin{center}
		\leavevmode
		\subfloat{
			\includegraphics[width=0.44\textwidth]{f6.pdf}
		}

		\vspace{-0.3cm}
		\subfloat{
			\includegraphics[width=0.47\textwidth]{f3.pdf}
		}
	\end{center}
	\vspace{-0.6cm}
	\caption{
    {\bf Speckle-imaging and velocimetry of \tn.} {\it Top:} SOAR
    HRCam contrast limits derived from point-source injection-recovery
    experiments. Star A ($\Delta T=4.7$, $2.3''$ West) is detected,
    and is also a resolved Gaia source.  It is co-moving with \tn, and
    its parallax and on-sky position imply that it is physically
    separated from \tn\ by $6.6\pm 0.1\,{\rm pc}$.  {\it Middle:}
    Radial velocity (RV) measurements, with best-fit instrument
    offsets and jitter terms included.  The expected scatter from
    starspots based on $v\sin i$ and the photometric modulation
    amplitude is of order 300m$\,$s$^{-1}$.  {\it Bottom:} RV
    measurements phased to the orbital ephemeris of \pn.  The planet
    is not detected.  The black line shows a circular Keplerian orbit
    representing the 3-$\sigma$ upper mass limit.
		\label{fig:followup}
	}
\end{figure}

@@

We derived radial velocities and spectroscopic line profiles from the
CHIRON observations using a least-squares deconvolution of the spectra
against non-rotating synthetic spectral templates
\citep{donati_1997_spectropolarimetric}. The spectral templates were
generated using ATLAS9 atmosphere models \citep{Castelli:2004} with
the \texttt{SPECTRUM} script \citep{gray_1994_spectrum}. These line
profiles were fitted with a broadening kernel that describes the
rotational, radial-tangential macroturbulent, and instrumental
broadening of the spectrum. The rotational and macroturbulent
broadening are computed per \citet{gray_2005_book}, following the
methods described in \citet{zhou_2018_hd106315obliq}. We fitted the
line profile from each observation independently, yielding the radial
velocities listed in Table~\ref{tab:rvs} and shown in
Figure~\ref{fig:followup}. We found a mean rotational broadening
velocity of $v\sin I_\star = 16.2 \pm 1.1 \mathrm{km\,s}^{-1}$, and a
macroturbulent broadening of $v_\mathrm{Mac} = 8.4 \pm 2.9
\mathrm{km\,s}^{-1}$.

@@

\subsubsection{FEROS}
TOI 837.01 was monitored with the FEROS echelle spectrograph
\citep{kaufer_commissioning_1999}, mounted on the MPG 2.2$\,$m
telescope at the ESO La Silla Observatory, in Chile. FEROS has a
resolution of $\approx$48,000 across a spectral range of 350–920 nm.
It has a remarkably high efficiency of $\approx$20\%. We obtained 13
spectra of \tn\ between July 5 of 2019 and March 14 of 2020 in the
context of the Warm gIaNts with tEss (WINE) collaboration, which
focuses on the systematic characterization of TESS transiting giant
planets with moderately long orbital periods \citep[{\it
e.g.},][]{brahm:2019,jordan:2020}.  We adopted exposure times of 500
and 600 seconds and the observations were performed with the
simultaneous calibration mode for tracing the instrumental velocity
variations with a comparison fiber illuminated with a ThAr lamp. FEROS
data was processed with the \texttt{ceres} pipeline
\citep{brahm_2017_ceres}, which delivers precision radial velocities
and bisector span measurements through cross-correlation of the
extracted spectra with a binary mask resembling the properties of a
G2V star. The radial velocities are given in Table~\ref{tab:rvs}, and
shown in Figure~\ref{fig:followup}.
To check for the presence of secondary lines, we performed a similar
injection-recovery exercise as with the CHIRON data. 
We achieved slightly worse limits, likely due to the lower spectral
resolution of FEROS, and therefore adopted the CHIRON limits.

\subsubsection{Veloce}
We acquired 34 spectra over 10 visits of \tn\ using the Veloce
spectrograph, mounted on the 3.9$\,$m Anglo-Australian Telescope at
Siding Spring Observatory near Coonabarabran, Australia
\citep{gilbert_veloce_2018}.  Veloce's currently implemented Rosso
camera provides coverage from 600--950$\,$nm at a spectral resolution
of $\approx 80{,}000$.  Many of the exposures were taken in average or
poor seeing conditions, when fiber-to-fiber cross-contamination on the
IFU-style fiber feed is strongest.  To reduce the spectra to
velocities, we cross-correlated against a high S/N template of
$\delta$ Pavonis, because with spectral type G8 IV it was the closest
\tn\ analog available in the Veloce spectral database.  The velocity
RMS seen across each visit was hundreds of meters per second, likely
due to uncorrected fiber-to-fiber cross-contamination.  For analysis
purposes, we averaged across each visit, and set the velocity
uncertainties to be the standard deviation of the per-visit exposures.
The velocities are given in Table~\ref{tab:rvs}, and shown in
Figure~\ref{fig:followup}.

@@

\begin{figure*}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.8\textwidth]{f4.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{
    {\bf Astrophysical false positive scenarios.} {\it Top}: for bound
    companions (EB and HEB scenarios), {\it Bottom}: for unassociated
    companions along the same line of sight (BEB scenarios).  Each
    constraint is described in Section~\ref{subsec:fp_constraints}.
    Gray regions are ruled out by at least one constraint.
		\label{fig:fpscenario}
	}
\end{figure*}

@@

\subsubsection{Transit Depth}
In HEB and BEB scenarios, the flux from \tn\ and the true eclipsing
binary host blend together, and reduce the ``true'' eclipse depth
$\delta_{\rm true}$ to the observed depth
$\delta_{\rm obs}$:
\begin{equation}
  \delta_{\rm obs}
  = 
  \delta_{\rm true} \frac{F_{\rm comp}}{F_{\rm total}},
\end{equation}
where the total system flux and the flux from only the companion
(``comp'') binary are labeled appropriately.  The requirement that the
eclipse is produced by fusion-powered stars and that $\delta_{\rm true} < 0.5$
translates to a bound on the faintest possible blended companion
system:
\begin{equation}
  \Delta m < -\frac{5}{2} \log_{10}
             \left( \frac{0.5}{\delta_{\rm obs}} \right).
\end{equation}
For \tn\ ($T=9.93$), this implies that any stellar companion invoked
to explain the transit depth must be brighter than $T=15.07$.  In
Figure~\ref{fig:fpscenario}, we set the spatial limit to 2$''$ based
on the precision at which we have localized the transits using
seeing-limited ground-based photometry.

@@

% timmy.drivers.oneoff_drivers.get_mass_upper_limit
For the first fit we set a prior on the period and
time of conjunction using the known ephemeris from the transit.  We then
fitted for the semi-amplitude, instrument offsets, and jitter
parameters.  This yielded a non-detection of the planet's orbit.  The
corresponding 3-$\sigma$ (99.7$^{\rm th}$ percentile) upper limit on
$M_{\rm p} \sin i$ is $1.20 M_{\rm Jup}$.  The data and corresponding
model are shown in Figure~\ref{fig:followup}.

@@

\subsection{False positive probability}

@@

\tn\ joins a number of other young planetary systems reported from
TESS, including DS~Tuc~Ab, HIP~67522b, TOI~1726, and AU~Mic~b
\citep{newton_tess_2019,zhou_well_2019,montet_young_2019,rizzuto_tess_2020,mann_tess_2020,plavchan_planet_2020,palle_transmission_2020,addison_youngest_2020,martioli_magnetism_2020,hirano_limits_2020}.
Figure~\ref{fig:rpvsage} shows \tn\ in the space of planet sizes and
ages.  \tn\ is among the youngest transiting planets known.  While we
have statistically validated that \tn\ is a planet, the possibility
that it is a background eclipsing binary has not been excluded with
sufficient confidence to call the planet ``confirmed''.

The easiest path towards definitively confirming whether \tn\ is a
planet will be a Rossiter-McLaughlin (RM) measurement.  Detection of
an RM signal consistent with the photometric transit would rule out
BEB and HEB scenarios, as it would imply that the eclipsing object is
bound to the target star. Combined with our non-detection of the
planet's mass from radial velocity monitoring, this would confirm that
\pn\ is a planet.

@@

The Rossiter-McLaughlin approach is more likely to yield short-term
success than a direct mass measurement because of the RV noise
expected to be induced by stellar rotation.  The photometric amplitude
induced by starspots on \tn\ is $\approx 2\%$.  The spot-induced RV
variation expected over the course of the $\approx 3\,{\rm d}$
rotation period can then be estimated by multiplying the photometric
amplitude and spectroscopic equatorial velocity.  This gives
$\sigma_{\rm RV,rot} \approx 300\,{\rm m\,s}^{-1}$, and is consistent
with the scatter we observe in our radial velocities from FEROS.
Detecting a planet's Doppler signal in these regimes is challenging,
and requires a significant amount of data and care in signal
extraction \citep{barragan_radial_2019}.  The Rossiter-McLaughlin
measurement avoids the majority of this issue because the transit
occurs over a much shorter duration than a single stellar rotation
period.

If the RM measurements prove that the validated planet is real, a
campaign to measure the mass timed to coincide with TESS sectors 36
and 37 (late March to late May of 2021) would significantly ease the
extraction of the planetary Doppler signal.  The reason is that the
RVs, activity indicators, and photometry could be modeled
simultaneously \citep[{\it e.g.},][]{rajpaul_gaussian_2015}. This
would greatly constrain models of the spot-induced RV signal.  The
mass measurement would likely be worthwhile in order to improve
understanding of the planet's contraction and photoevaporation
history.

While we hope that RV observations will be pursued, data acquired
during the first TESS mission extension may also help in understanding
the system \citep{bouma_extend_2017,huang_expected_2018}.  Any
misalignment between the star's spin axis and the planet's orbit
should induce nodal precession, which could yield large changes in the
transit duration given the system's high impact parameter.  If
observed, this could be a spectroscopy-free method for confirming the
planet.

>>>>

The observations of \tn\ were initially acquired by
\citet{ziegler_soar_2020} as part of the Southern Astrophysical
Research (SOAR) TESS Survey using the High Resolution Camera (HRCam;
\citealt{tokovinin_ten_2018}).  The HRCam $I$-band filter is described
by \citet{tokovinin_ten_2018}.  The points in
Figure~\ref{fig:speckle} show the resulting measured 5-$\sigma$
detectable contrasts.  The lines are linear smoothing fits between the
regimes of the diffraction limit, the ``knee'' at $\approx 0.2''$, and
the slow decrease until $\approx 1.5''$, beyond which the speckle
patterns become de-correlated.  Star A (TIC 847769574) was detected at
the expected location and brightness contrast, and no additional
companions were found.  Star B was not detected; with a separation of
$5.4''$ from \tn, it fell outside the field of view.

@@

\begin{figure}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=.44\textwidth]{f6.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{
    {\bf Speckle-imaging of \tn.} Contrast limits from SOAR
    HRCam imaging were derived from point-source injection-recovery
    experiments. Star A ($\Delta T=4.7$, $2.3''$ West) is detected,
    and is also a resolved Gaia source.  It is co-moving with \tn, and
    its parallax and on-sky position imply that it is physically
    separated from \tn\ by $6.6\pm 0.1\,{\rm pc}$.  
		\label{fig:speckle}
	}
\end{figure}

\begin{figure}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=.47\textwidth]{f3.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{
    {\bf Velocimetry of \tn.} {\it Top:} Radial velocity (RV)
    measurements, with best-fit instrument offsets and jitter terms
    included.  The expected scatter from starspots based on $v\sin i$
    and the photometric modulation amplitude is of order
    300m$\,$s$^{-1}$.  {\it Bottom:} RV measurements phased to the
    orbital ephemeris of \pn.  The planet is not detected.  The black
    line shows a circular Keplerian orbit representing the 3-$\sigma$
    upper mass limit.
    \label{fig:rvs}
	}
\end{figure}

@@

We derived radial velocities and spectroscopic line profiles from the
CHIRON observations using a least-squares deconvolution of the spectra
against non-rotating synthetic spectral templates
\citep{donati_1997_spectropolarimetric}. The spectral templates were
generated using ATLAS9 atmosphere models \citep{Castelli:2004} with
the \texttt{SPECTRUM} script \citep{gray_1994_spectrum}. These line
profiles were fitted with a broadening kernel that describes the
rotational, radial-tangential macroturbulent, and instrumental
broadening of the spectrum. The rotational and macroturbulent
broadening are computed per \citet{gray_2005_book}, following the
methods described in \citet{zhou_2018_hd106315obliq}. We fitted the
line profile from each observation independently, yielding the radial
velocities listed in Table~\ref{tab:rvs} and shown in
Figure~\ref{fig:rvs}. We found a mean rotational broadening
velocity of $v\sin I_\star = 16.2 \pm 1.1 \mathrm{km\,s}^{-1}$, and a
macroturbulent broadening of $v_\mathrm{Mac} = 8.4 \pm 2.9
\mathrm{km\,s}^{-1}$.

@@

\subsubsection{FEROS}
TOI 837.01 was monitored with the FEROS echelle spectrograph
\citep{kaufer_commissioning_1999}, mounted on the MPG 2.2$\,$m
telescope at the ESO La Silla Observatory, in Chile. FEROS has a
resolution of $\approx$48,000 across a spectral range of 350–920 nm.
It has a remarkably high efficiency of $\approx$20\%. We obtained 13
spectra of \tn\ between July 5 of 2019 and March 14 of 2020 in the
context of the Warm gIaNts with tEss (WINE) collaboration, which
focuses on the systematic characterization of TESS transiting giant
planets with moderately long orbital periods \citep[{\it
e.g.},][]{brahm:2019,jordan:2020}.  We adopted exposure times of 500
and 600 seconds and the observations were performed with the
simultaneous calibration mode for tracing the instrumental velocity
variations with a comparison fiber illuminated with a ThAr lamp. FEROS
data was processed with the \texttt{ceres} pipeline
\citep{brahm_2017_ceres}, which delivers precision radial velocities
and bisector span measurements through cross-correlation of the
extracted spectra with a binary mask resembling the properties of a
G2V star. The radial velocities are given in Table~\ref{tab:rvs}, and
shown in Figure~\ref{fig:rvs}.
To check for the presence of secondary lines, we performed a similar
injection-recovery exercise as with the CHIRON data. 
We achieved slightly worse limits, likely due to the lower spectral
resolution of FEROS, and therefore adopted the CHIRON limits.

\subsubsection{Veloce}
We acquired 34 spectra over 10 visits of \tn\ using the Veloce
spectrograph, mounted on the 3.9$\,$m Anglo-Australian Telescope at
Siding Spring Observatory near Coonabarabran, Australia
\citep{gilbert_veloce_2018}.  Veloce's currently implemented Rosso
camera provides coverage from 600--950$\,$nm at a spectral resolution
of $\approx 80{,}000$.  Many of the exposures were taken in average or
poor seeing conditions, when fiber-to-fiber cross-contamination on the
IFU-style fiber feed is strongest.  To reduce the spectra to
velocities, we cross-correlated against a high S/N template of
$\delta$ Pavonis, because with spectral type G8 IV it was the closest
\tn\ analog available in the Veloce spectral database.  The velocity
RMS seen across each visit was hundreds of meters per second, likely
due to uncorrected fiber-to-fiber cross-contamination.  For analysis
purposes, we averaged across each visit, and set the velocity
uncertainties to be the standard deviation of the per-visit exposures.
The velocities are given in Table~\ref{tab:rvs}, and shown in
Figure~\ref{fig:rvs}.

@@

\begin{figure*}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.75\textwidth]{f4.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{
    {\bf Astrophysical false positive scenarios.} {\it Top}: for bound
    companions (EB and HEB scenarios), {\it Bottom}: for unassociated
    companions along the same line of sight (BEB scenarios).  Each
    constraint is described in Section~\ref{subsec:fp_constraints}.
    Gray regions are ruled out by at least one constraint.
		\label{fig:fpscenario}
	}
\end{figure*}

@@

\subsubsection{Transit Depth}
In HEB and BEB scenarios, the flux from \tn\ and the true eclipsing
binary host blend together, and reduce the ``true'' TESS-band eclipse depth
$\delta_{\rm true}$ to the observed depth
$\delta_{\rm obs}$:
\begin{equation}
  \delta_{\rm obs}
  = 
  \delta_{\rm true} \frac{F_{\rm comp}}{F_{\rm total}},
\end{equation}
where the total system flux and the flux from only the companion
(``comp'') binary are labeled appropriately.  The requirement that the
eclipse is produced by fusion-powered stars and that $\delta_{\rm true} < 0.5$
translates to a bound on the faintest possible blended companion
system:
\begin{equation}
  \Delta m < -\frac{5}{2} \log_{10}
             \left( \frac{0.5}{\delta_{\rm obs}} \right).
\end{equation}
For \tn\ ($T=9.93$), this implies that any stellar companion invoked
to explain the transit depth must be brighter than $T=15.07$.  In
Figure~\ref{fig:fpscenario}, we set the spatial limit to 2$''$ based
on the precision at which we have localized the transits using
seeing-limited ground-based photometry.

@@

% timmy.drivers.oneoff_drivers.get_mass_upper_limit
For the first fit we set a prior on the period and
time of conjunction using the known ephemeris from the transit.  We then
fitted for the semi-amplitude, instrument offsets, and jitter
parameters.  This yielded a non-detection of the planet's orbit.  The
corresponding 3-$\sigma$ (99.7$^{\rm th}$ percentile) upper limit on
$M_{\rm p} \sin i$ is $1.20 M_{\rm Jup}$.  The data and corresponding
model are shown in Figure~\ref{fig:rvs}.

@@

\subsection{False positive probability}
\label{subsec:fpp}

@@

\tn\ joins a number of other young planetary systems reported from
TESS, including DS~Tuc~Ab, HIP~67522b, TOI~1726, and AU~Mic~b
\citep{newton_tess_2019,zhou_well_2019,montet_young_2019,rizzuto_tess_2020,mann_tess_2020,plavchan_planet_2020,palle_transmission_2020,addison_youngest_2020,martioli_magnetism_2020,hirano_limits_2020}.
Figure~\ref{fig:rpvsage} shows \tn\ in the space of planet sizes and
ages.  \tn\ is among the youngest transiting planets known.

While we
have statistically validated that \tn\ is a planet,
the possibility
that it could be a background eclipsing binary has not been excluded with
sufficient confidence to call the planet ``confirmed''.
The distinction is methodological.
Our calculations have shown that at a population
level we expect negligibly few BEBs within $\approx0.3''$ of \tn\ to
produce eclipses of the appropriate shape across all bandpasses, with
no observed secondary eclipse or odd-even variations (Section~\ref{subsec:fpp}).
This statement is tautologically ``validation'', but it is weaker
than having data on hand that conclusively
rules in favor of the planetary interpretation.

The easiest path towards definitively confirming the planetary
nature of \tn\ will be a Rossiter-McLaughlin (RM) measurement.  Detection of
an RM signal consistent with the photometric transit would rule out
BEB and HEB scenarios, as it would imply that the eclipsing object is
bound to the target star. Combined with our non-detection of the
planet's mass from radial velocity monitoring, this would confirm that
\pn\ is a planet.

@@

The Rossiter-McLaughlin approach is more likely to yield short-term
success than a direct mass measurement because of the RV noise
expected to be induced by stellar rotation.  The photometric amplitude
induced by starspots on \tn\ is $\approx 2\%$.  The spot-induced RV
variation expected over the course of the $\approx 3\,{\rm d}$
rotation period can be estimated by multiplying the photometric
amplitude and spectroscopic equatorial velocity.  This gives
$\sigma_{\rm RV,rot} \approx 300\,{\rm m\,s}^{-1}$, and is consistent
with the scatter we observe in our radial velocities from FEROS.
Detecting the radial velocity signature of a planet's Keplerian motion
in this regime is challenging, and requires a significant amount of
data and care in signal extraction \citep{barragan_radial_2019}.  The
Rossiter-McLaughlin measurement avoids the majority of this issue
because the transit occurs over a much shorter duration than a single
stellar rotation period.

If the RM measurements prove that the validated planet is real, a
campaign to measure the mass timed to coincide with TESS sectors 36
and 37 (late March to late May of 2021) would significantly ease the
extraction of the planetary mass.  The reason is that the RVs,
activity indicators, and photometry could be modeled simultaneously
\citep[{\it e.g.},][]{rajpaul_gaussian_2015}. This would greatly
constrain models of the spot-induced RV signal.  The mass measurement
would be a worthy goal in order to improve understanding of the
planet's contraction and photoevaporation history.

While we hope that RV observations will be pursued, data acquired
during the TESS mission extension may also help in understanding
the system \citep{bouma_extend_2017,huang_expected_2018}.  Any
misalignment between the star's spin axis and the planet's orbit
should induce nodal precession, which could yield large changes in the
transit duration given the system's high impact parameter.  If
observed, this could be a spectroscopy-free method for confirming the
planet.
